---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# wbstats: An R package for searching and downloading data from the World Bank API.

You can install:

The latest release version from CRAN with
```{r, eval = FALSE}
install.packages("wbstats")
```

or

The latest development version from github with
```{r, eval = FALSE}
remotes::install_github("nset-ornl/wbstats")
```

# Introduction

The World Bank^[<http://www.worldbank.org/>] is a tremendous source of global socio-economic data; spanning several decades and dozens of topics, it has the potential to shed light on numerous global issues. To help provide access to this rich source of information, The World Bank themselves, provide a well structured RESTful API^[<http://data.worldbank.org/developers>]. While this API is very useful for integration into web services and other high-level applications, it becomes quickly overwhelming for researchers who have neither the time nor the expertise to develop software to interface with the API. This leaves the researcher to rely on manual bulk downloads of spreadsheets of the data they are interested in. This too is can quickly become overwhelming, as the work is manual, time consuming, and not easily reproducible. The goal of the `wbstats` R-package is to provide a bridge between these alternatives and allow researchers to focus on their research questions and not the question of accessing the data. The `wbstats` R-package allows researchers to quickly search and download the data of their particular interest in a programmatic and reproducible fashion; this facilitates a seamless integration into their workflow and allows analysis to be quickly rerun on different areas of interest and with realtime access to the latest available data.

```{r readme-chart, warning=FALSE, message=FALSE, fig.width=10, fig.height=6}
library(tidyverse)
library(wbstats)

wb_data(
  c(life_exp = "SP.DYN.LE00.IN", 
    gdp_capita ="NY.GDP.PCAP.CD", 
    pop = "SP.POP.TOTL"), 
  start_date = 2016
  ) %>%
  left_join(wb_countries(), "iso3c") %>%
  ggplot() +
  geom_point(
    aes(
      x = gdp_capita, 
      y = life_exp, 
      size = pop, 
      color = region
      )
    ) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    breaks = scales::log_breaks(n = 10)
    ) +
  coord_trans(x = 'log10') +
  scale_size_continuous(
    labels = scales::number_format(scale = 1/1e6, suffix = "m"),
    breaks = seq(1e8,1e9, 2e8),
    range = c(1,20)
    ) +
  theme_minimal() +
  labs(
    title = "An Example of Hans Rosling's Gapminder using wbstats",
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy at Birth",
    size = "Population",
    color = NULL,
    caption = "Source: World Bank"
  ) 
    
```
